Computing likelihoods under the SMC for a general class of ARGs.

### 2024-04-25 paper.tex
update process description

\begin{figure}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG (top), generated under the SMC, and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows show the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
By removing recombinant nodes, we limit the information on the time to each
of these events to the windows (blue shaded array) delimited by by the age of
the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

@@

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} genomic intervals may coalesce.
By limiting common ancestry events to overlapping pairs of lineages, we are
guaranteed that all nodes in the ARG only carry ancestral material, this is
genetic material that is inherited by either one of a set of samples \citep{wiuf_ancestry_1999}.
As such, we only require a single half-open interval to describe the ancestral
material associated with each common ancestor node in the ARG.
Surprisingly, although the SMC can be defined
in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.
Linking this back to relationship between a graph and local trees view on
the ARG, then the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
(To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.

@@

To ensure that the ancestral material carried by a single lineage can at
all time be represented by a single half-open interval, we only update the
endpoints of the involved lineages following common ancestry or recombination
such that only the minimal half-open interval whose endpoints have not fully coalesced
is used to update $L(t)$. Note that empty intervals are not added to the state.
%By reasoning about the process in this way we can keep representing 
% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% and the process terminates.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    \caption{
        Graphical representation of the events associated with an edge $e(p,c,x_c,y)$.
        The lineage carried by $c$ spans $[x,y)$ until a recombination event splits it
        into two segments. The right hand segment coalesces at time $t_p$. The time to
        recombination is thus bounded by $(c. \min(t_p, t_{p'})$.
        \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$.
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times S)$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occurr, then the number of links not
affected by recombination is 1 less (see \ref{eq:depth}).

>>>>

\begin{figure}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. By removing recombinant nodes, we limit the information on the time to each
of these events to the windows (blue shaded array) delimited by by the age of
the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
Blue (unary) nodes represent recombinant nodes. The blue arrows show the
familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
}
\label{fig:smc-unary}
\end{figure}

@@

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} genomic intervals may coalesce.
By limiting common ancestry events to overlapping pairs of lineages, we are
guaranteed that all nodes in the ARG only carry ancestral material, this is,
genetic material that is inherited by either one of a set of samples \citep{wiuf_ancestry_1999}.
As such, we only require a single half-open interval to describe the ancestral
material associated with each common ancestor node in the ARG.
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
Indeed, the resulting tree-valued process along the genome is also Markov.
Linking this back to the relationship between the ``graph'' and
``local trees'' representation of
the ARG, this can be formulated as follows: the distribution of the
next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be a coalescent node or it already has been:
looking at the second tree (Fig.\ref{fig:smc-unary}),
we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.

@@

To ensure that the ancestral material carried by a single lineage can at
all time be represented by a single half-open interval, we only update the
endpoints of the involved lineages following common ancestry or recombination events
such that only the minimal half-open interval whose endpoints have not fully coalesced
is used to update $L(t)$. Note that empty intervals are not added to the state.
%By reasoning about the process in this way we can keep representing 
% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% and the process terminates.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    \caption{
        Graphical representation of the events associated with an edge $(p,c,x,y)$.
        The lineage carried by $c$ spans $[x',y)$ until a recombination event splits it
        into two segments. The right hand segment coalesces at time $t_p$. The time to
        recombination is thus bounded by $(c. \min(t_p, t_{p'})$.
        \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$.
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occurr, then the number of links not
affected by recombination is 1 less (see \ref{eq:depth}).
